Fail Faster

### 2025-03-19 paper/paper.tex
2.1 finished ish

While many DSL designs are possible, the standard design for QuickCheck-style
generators is via an embedded {\em monadic} language. For the purposes of this
paper, we use the sytnax from \bq, which is presented in
Figure~\ref{fig:bq-api}. The library provides a series of basic generators, for
example ones like \texttt{gen\_int} for random integers in a range, along with
the ``monadic interface'' consisting of \texttt{return} and \texttt{bind}.  The
generator \texttt{return x} is the constant generator, always generating the
value \texttt{x}. Running a generator \texttt{bind g k} runs the generator
\texttt{g}, producing a value \texttt{a}, and then runs the generator \texttt{k
a}.
\iflater
\jwc{\texttt{gen\_int} here is really \texttt{int\_uniform\_inclusive}. I also
think we should completely omit named arguments in the paper so we don't confuse
with splices.}
\fi

Together, these three functions are the bare minimum for constructing arbitrary
random data generators: \texttt{gen\_int} provides a base source of randomness,
and \texttt{return} and \texttt{bind} allow generators to be composed to create
larger, more complex generators. Figure~\ref{fig:simple-bind} shows a generator
built with these operations; it first samples an int between $0$ and $100$,
names it $x$, samples another between $0$ an $x$, and then returns the pair of
them both. In practice, PBT generator libraries also include other functions
that make generator construction easier, including functions like
\texttt{weighted\_union} for making a weighted choice between different
generators as well as functions for managing the sizes of generated values and
combinators for recursion.

@@

Most languages in which monadic APIs are common expose some sort of syntactic
sugar for them. In OCaml\footnote{OCaml actually has a few versions of syntax
for monadic APIs; this is the one provided by Jane Street's libraries.}, this
looks like \texttt{let\%bind x = e in e'}, which is a macro that desugars to
\texttt{bind e (fun x => e')}. Figure~\ref{fig:simple-macro} shows the same
generator as before, now written with macros.

\jwc{
Explain effect ordering here!!
\begin{itemize}
  \item \texttt{Bq.bind} has the CBV sampling semantics
  \item If you \texttt{Bq.bind} \texttt{x} and then use it twice, both uses refer to the \emph{same} value.
  \item Sampling is effectful: it mutates the internal state.
  \item Permuting two independent binds gives you a generator that is distributionally equivalent, but not equivlaent when considered up to pointwise equality as \emph{functions} from seed to value.
  \item Note that it's much more sensible to compare the bugfinding performance of two function-equivalent generators:
\end{itemize}
}

\jwc{
  \begin{itemize}
    \item PBT generator libraries include more functions than just the monad interface and sampling, libraries include helper functions.
    \item With \texttt{weighted\_union}, to make a weighted choices, and \texttt{fixed\_point} to define recursive generators.
    \item Also \texttt{size} and \texttt{with\_size} to adjust the size parameter --- these are useful to ensure we terminate with recursive generators.
  \end{itemize}
}

\jwc{This section needs to explain approximately how weighted union works --- compute a sum of the weights, sample between 0 and the sum, pick
the first element where the partial sums to the left exceeds the sampled number.
(This has been written about a billion times, cite one of the papers about this.)
}
\jwc{I re-explain this later somewhere i think, so we need to remember to RM that explanation.}

\begin{lstlisting}
let tree_of g = fixed_point (fun rg ->
  let%bind n = size in
  weighted_union [
    (1, return E);
    (n,
      let%bind x = g in
      let%bind l = with_size (n / 2) rg in
      let%bind r = with_size (n / 2) rg in
      return (Node (l,x,r))
    )
  ]
)
\end{lstlisting}

\jwc{This generator generates binary trees using all the combinators, yay. Explain em.}

\jwc{IMPORTANT: We are using OCaml because we have to pick a syntax for this section, but these abstraction overheads exist in other languages -- at least scala, rust.}

>>>>

While many DSL designs are possible, the standard design for QuickCheck-style
generators is via an embedded {\em monadic} language. For the purposes of this
paper, we use the syntax from \bq, which is presented in
Figure~\ref{fig:bq-api}, but these kinds of DSLs can also be found in languages
like Haskell, Scala, Rust, and many more.
The library provides some basic generators, for
example \texttt{gen\_int} for generating random integers in a range, along with
the ``monadic interface'' consisting of \texttt{return} and \texttt{bind}.  The
generator \texttt{return x} is the constant generator, always generating the
value \texttt{x}. Running a generator \texttt{bind g k} runs the generator
\texttt{g}, producing a value \texttt{a}, and then runs the generator \texttt{k
a}.
\iflater
\jwc{\texttt{gen\_int} here is really \texttt{int\_uniform\_inclusive}. I also
think we should completely omit named arguments in the paper so we don't confuse
with splices.}
\fi

Together, these three functions are the bare minimum for constructing arbitrary
random data generators: \texttt{gen\_int} provides a base source of randomness,
and \texttt{return} and \texttt{bind} allow generators to be composed to create
larger, more complex generators. Figure~\ref{fig:simple-bind} shows a generator
built with these operations; it first samples an int between $0$ and $100$,
names it $x$, samples another between $0$ an $x$, and then returns the pair of
them both.

@@

Most languages in which monadic APIs are common expose some sort of syntactic
sugar for them. In OCaml\footnote{OCaml actually has a few ways to implement
monadic syntax; this is the one provided by Jane Street's libraries.}, this
looks like \texttt{let\%bind x = e in e'}, which is a macro that desugars to
\texttt{bind e (fun x => e')}. Figure~\ref{fig:simple-macro} shows the same
generator as before, now written with macros.

\subsubsection{Subtleties of Bind}
The \texttt{bind} function in monadic generator libraries sequences generators
with a call-by-value semantics, regardless of the calling convention of the
language in which it is implemented. Concretely, this means that sampling
\begin{center}
  \texttt{let\%bind x = gen\_int 0 10 in return (x + x)}
\end{center}
always produces an even number; \texttt{x} is only sampled once (when the
\texttt{bind} is evaluated) and from that point on it is a pure value. This is
in contrast to the semantics of some probabilistic programming languages that
may have call-by-name binders. This is important because \hg{Joe why are we
talking about this here? What are we going to use this fact for?}.

Another subtle aspect of \texttt{bind} is that many implementations mutate the
internal state of the random number generator\footnote{The original QuickCheck
implementation in Haskell uses a splittable random number generator~\cn, and
\texttt{bind} splits the seed.} This means that generators do not always obey
equivalences that one might expect from a monadic program. For example,
\begin{center}
  \texttt{let\%bind \_x = return () in gen\_int 0 10}
\end{center}
does not necessarily produce the same sequence of values as \texttt{gen\_int 0
10}; the \texttt{bind} updates the random seed in the background before
geenrating the integer, even though it looks like a no-op. Critically, while
these generators do not produce the same sequence of values, they do produce the
same {\em distributions} of values. We say that generators obey standard monadic
equalities up to ``distributional equivalence.''
In Section~\hg{todo}, we discuss why distributional equivalence is often weaker
than one might like when evaluating the differences between
generators.\hg{Update this}

\subsubsection{Weighted Union and Other Conveniences}
Generator libraries also include other functions that make generator
construction easier; some examples of these are also shown in
Figure~\ref{fig:bq-api}.  The \texttt{weighted\_union} function is a
particularly well-used one: it makes making a weighted choice between different
generators, allowing the developer to combine different sub-generators into a
single program and tune the data distribution. Also important are functions like
\texttt{size} and \texttt{with\_size} that are used to control the sizes of
generated values and \texttt{fixed\_point} that is used tto define recursive
generators.

\begin{wrapfigure}{r}{.5\textwidth}
\begin{lstlisting}
let tree_of g = fixed_point (fun rg ->
  let%bind n = size in
  weighted_union [
    (1, return E);
    (n,
      let%bind x = g in
      let%bind l = with_size (n / 2) rg in
      let%bind r = with_size (n / 2) rg in
      return (Node (l,x,r)))
  ])
\end{lstlisting}
\caption{A generator using a variety of convenience functions.}\label{fig:tree_of}
\end{wrapfigure}

The generator in Figure~\ref{fig:tree_of} uses all of the aforementioned
features. It uses \texttt{fixed\_point} to define a recursive generator that
reads the current value of \texttt{size} to determine how to generate a tree.
The generator uses \texttt{weighted\_union} to make a random choice between an
empty tree and a node, choosing a node with weight proportional to the current
size and choosing a leaf otherwise. When generating a node, the generator uses
\texttt{with\_size} to reduce the value of the size parameter for future
iterations.
